---
title: "Introduction to dplyr"
author: "Jason"
date: "2/14/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install install dplyr
# install.packages("dplyr", deps=T)

library(dplyr)
```

## dplyr verbs
- select: pick variables based on column names.
- filter: pick variables based on row values.
- mutate: allows us to create new columns based on existing data
- arrange: allows us to sort/order data variables
- summarise: reduce values to a single summary
- group_by: allows us to work on dataset in groups/categories.

## Load data

```{r load.data}
crop.prod.df <- read.csv("data/cropprodandvalue_2015to19_tb1.csv")
crop.prod.df
```
### Prepare dataset
a vector header

```{r header.col}
header <- c(
  "crop",
  "prod.MT.2015",
  "value.ksh.2015",
  "prod.MT.2016",
  "value.ksh.2016",
  "prod.MT.2017",
  "value.ksh.2017",
  "prod.MT.2018",
  "value.ksh.2018",
  "prod.MT.2019",
  "value.ksh.2019"
)
names(crop.prod.df) <- header
crop.prod.df
```

```{r slice.rows}
crop.prod.df[c(1, 2, 3, 4),]

crop.prod.df[c(-1, -2, -3, -4), ]

crop.prod.df <- crop.prod.df[c(-1, -2), ]
crop.prod.df
```

```{r subcounty.col}

crop.prod.df[2:8, "subcounty"] <- "Mbooni"
crop.prod.df[10:16, "subcounty"] <- "Kilome"
crop.prod.df[18:24, "subcounty"]  <- "Kaiti"

crop.prod.df[, c("crop","subcounty")]


```

### dplyr verbs

#### filter
allows us to pick rows based on cell value

```{r filter}
filter(crop.prod.df, is.na(subcounty))

crop.prod.df <- filter(crop.prod.df, !is.na(subcounty))
crop.prod.df
```

#### relocate
shift column positions

```{r relocate}
relocate(crop.prod.df, .before = 1, subcounty)

crop.prod.df <- relocate(crop.prod.df, .after=1, subcounty)
crop.prod.df
```
```{r chaining}
crop.prod.df <- crop.prod.df %>% relocate(.before=1, subcounty)

crop.prod.df
```

````{r num.conversion}
val = "36,016"

as.numeric( gsub(",", "", val))

#mutate(crop.prod.df, prod.MT.2015 = gsub(",", "", prod.MT.2015))

#mutate(crop.prod.df, prod.MT.2015 = as.numeric(gsub(",", "", prod.MT.2015)))

to_numeric <- function(value){
  as.numeric(gsub(",", "", value))
}

crop.prod.df.clean <- crop.prod.df %>%
  filter_all(all_vars(. != "-")) %>%
  mutate(
    prod.MT.2015 = to_numeric(prod.MT.2015),
    value.ksh.2015 = to_numeric(value.ksh.2015),
    prod.MT.2016 = to_numeric(prod.MT.2016),
    value.ksh.2016 = to_numeric(value.ksh.2016),
    prod.MT.2017 = to_numeric(prod.MT.2017),
    value.ksh.2017 = to_numeric(value.ksh.2017),
    prod.MT.2018 = to_numeric(prod.MT.2018),
    value.ksh.2018 = to_numeric(value.ksh.2018),
    prod.MT.2019 = to_numeric(prod.MT.2019),
    value.ksh.2019 = to_numeric(value.ksh.2019)
  )

crop.prod.df.clean
````

#### arrange
sort/order column values
```{r sort}
# ascending
crop.prod.df.clean %>% arrange(prod.MT.2015)

# descending
crop.prod.df.clean %>% arrange(desc(prod.MT.2015))
```




