---
title: "Introduction to R"
author: "clinton"
date: "2/14/2022"
output: html_document
---

R is an open source statistical programming language that is used for data wrangling, visualization and modelling. 
Its an object oriented programming language.   

## Basic arithmetic operations

### Addition
```{r}
# use (#) to comment 
# addition
6 + 8

```

### multiplication 
```{r}
9 * 5
```


### Variables  
Think of it like some sort of a container whose contents can change.    

```{r}
restaurant <- 87 # we use the assignment operator ( <- ) to assign values to variables
restaurant # what's in our variable? 

```

You can change the contents of a variable 

```{r}
restaurant <- 56
restaurant 
```


You can actually use objects/variables to perform arithmetic operations. For, imagine a vector with 1000 values, calling all that in order to manipulate the values is quite a task.  

```{r}

chefs <- 20

all_chefs <- restaurant * chefs
all_chefs
```

### Data structures  
A particular way of organizing data in a computer so that it can be used effectively.

#### Vector  
Collection of elements mostly character, logical, integer or numeric.  

```{r}
# when dealing with more than one value we use (c) to combine the values into a vector. 
stud_name <- c("Jack","Geddy","Jermin")
stud_name
```

#### list    
Collection of elements mostly of different data types.  

```{r}
makueni <- list(c("Agriculture","Education","FInance"),c(234,345,356))
makueni
```

#### dataframe     
A special type of list where every element of the list has same length (think of it as a rectangular list).   

```{r}
Makueni_subcounties <- data.frame(
  subcounties = c("Makueni","Kaiti","Kilome"),
  population = c(123,662453,735)
)
Makueni_subcounties
```

#### matrix    
A set of numbers arranged in rows and columns so as to form a rectangular array.  
Elements of a matrix must be of the same data type.   

```{r}
my_matrix <- matrix(c(1,2,3,4,5,6),ncol = 2,byrow = FALSE)
my_matrix
```
 
#### Factors  

In R, factors are used to work with categorical variables, variables that have a fixed and known set of possible values

```{r}
rating <- c( "low","low","high","low","high","medium","high")

the_levels <- c("low", "medium", "high")

rating_ordered <- factor(rating,levels = the_levels)

```

#### Data types
##### Character  

```{r}
first_name <- "Clinton"
first_name 
typeof(first_name)
```
 
##### numeric
```{r}
typeof(restaurant)
```
 
##### complex
```{r}
complex_n <-3 + 4i
complex_n
typeof(complex_n)
```
 
##### logical data type
```{r}
yes <- TRUE
typeof(yes)
```
 
 
 ## Assignment    
 
 1. Create a vector of counties 001 to 010, put the names into an object/variable and name it counties.       
 2. Create a vector of the populations of the above counties. Ensure the numbers align with the names of the counties. Put it into an object.variable called populations.     
 3. Using the above created objects, create a dataframe.    
 4. Create a factor object with all the months of the year arranged in order.     
 