---
title: "Introduction to mapping with leaflet"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Leaflet 

leaflet is an open-source library used to create interactive maps. 
With this library you can visualize your geospatial data easily. Given that the maps generated are interactive, we can embed them in R Markdown documents or Shiny web applications. 

## Installation and loading libraries
```{r}
# To install it use: 
#install.packages("leaflet") 
#install.packages("sf")

# To load the library into R session
library(leaflet) # for creating maps and visualizing geospatial data
library(sf) # for reading and manipulating vector data

```


# Brief about Spatial data

Spatial data: data with spatial reference i.e they have coordinate values and a system of reference for the coordinates.

There are two primary types of spatial data: 
  - raster data
  - vector data

* Raster data: a type of spatial data that consists of a matrix of cells organized into rows and columns in which each cell represents specific information i.e temperature, pH, elevation etc.

* Vector data: this is data that consists of sets of coordinate pairs (x , y) which are further categorized into 
  - point (individual x, y locations i.e location of Khartoum university library)
  - line (points that are connected i.e river)
  - polygon (set of closed polylines i.e Sudan boundaries)


# Creating a basic map 
```{r}

hosea <- leaflet() %>%
  setView(lng = 31.44, lat = -27.18, zoom = 9) %>%
  addTiles() %>% 
  addMarkers(lng = 31.4405, lat = -27.1825, popup = "Some place in Hosea") %>%
  addMiniMap()

hosea
```


## importing spatial data into R
*st_read()* function from *sf* library reads in the shapefiles into R. 
*read.csv* base function (comes with R) is used to read in .csv files. 

The type of data you can visualize on leaflet could be in the following formats: 
* .csv   
* .shp   
* .json   
* .tif   

As long as it has location details

## Working with .csv files.  

```{r} 

# import the csv file into R session   
ldsf.plots <- read.csv("./data/ldsf_plotproperties.csv")   

# inspect the data 
str(ldsf.plots) 

```

Map the csv file using the Longitude and Latitude location details. 

```{r}
ldsf.sites <- leaflet() %>%
  setView(lng = 31.4629694, lat = -26.5179414, zoom = 8) %>%
  addTiles() %>% 
  addMarkers(data = ldsf.plots,lng = ~Longitude, lat = ~Latitude) 

ldsf.sites
```


## Add state boundaries 
```{r}
# read in the shapefile with state boundaries. 
eswade.boundaries <- st_read("./data/shp/geoBoundaries-SWZ-ADM0.shp")

#summary
str(eswade.boundaries)
```


```{r}
ldsf.sites <- leaflet() %>%
  setView(lng = 31.4629694, lat = -26.5179414, zoom = 8) %>%
  addTiles() %>% 
  addMarkers(data = ldsf.plots,lng = ~Longitude, lat = ~Latitude) %>%
  addPolygons(data = eswade.boundaries)

ldsf.sites
```

## Add minimap 
```{r}
ldsf.sites <- leaflet() %>%
  setView(lng = 31.4629694, lat = -26.5179414, zoom = 8) %>%
  addTiles() %>% 
  addMarkers(data = ldsf.plots,lng = ~Longitude, lat = ~Latitude) %>%
  addPolygons(data = eswade.boundaries) %>% 
  addMiniMap()

ldsf.sites
```

## Interactive sessions on the browser (R Shiny context) 
```{r} 

ldsf.sites <- leaflet() %>%
  setView(lng = 31.4629694, lat = -26.5179414, zoom = 8) %>%
  addTiles() %>% 
  addMarkers(data = ldsf.plots,
             lng = ~Longitude, 
             lat = ~Latitude, 
             popup = paste("Site: ",ldsf.plots$Site)) %>%
  addPolygons(data = eswade.boundaries) %>% 
  addMiniMap()

ldsf.sites

```

